A Hedonic Price Analysis of Traffic Noise in the Twin Cities Housing Market

### 2013-06-24 reports/Draft.Rnw
continued working on draft of paper

\title{A Hedonic Price Analysis of Traffic Noise in the Twin Cities Housing Market}
\begin{document}
\maketitle

@@

\section{Past Research}\label{sec:lit}
Given the public policy ramifications of the issue, it is not surprising that multiple reviews of the economic impacts of traffic noise exist. \citet{Nelson1982} is perhaps the best known earliest review of 14 studies in Canada and the United States. More recently, a number of reviews have been produced for European government agencies. \citet{Bateman2001} reviewed the literature for the Scottish Government while \citet{Navrud2002} provided a ``state of the art'' review for the European Commission Environment Directorate General.

Since no explicit private market for noise abatement exists, the value of noise pollution is typically estimated through simulated or related markets. The Hedonic Price Method is commonly used to value noise by examining real estate price fluctuations as a function of noise levels. The estimation of the noise impact in the literature is typically reported as a percentage decrease in house price per one decibal (dB) increase in local traffic noise. According to \citet{Batemen2001}, the negative effect of traffic noise has ranged from 0.08\% to 2.22\% per dB with a mean around 0.55. \citet{Navrud2002} found an average decrease of 0.64\% per dB in a review of 65 noise evaluation studies. \citet{Nelson2008} produced an updated review of the traffic noise literature for the edited volume \underline{Hedonic Methods in Housing Markets Economics}. Nelson describes several areas of interest in the empirical literature, most notably: spatial modeling of the hedonic price function, long-run vs. short-run housing dynamics, and the appropriate measure of traffic noise. We now explain each of these areas in turn and our work's contribution to the field.

\subsection{Spatial Analysis}
Advances in Geographic Information System (GIS) technology and digitized datasets have made research easier to conduct over larger geographies, with larger samples, and while explicitly considering the spatial nature of the housing market. Ordinarly Least Squares estimation of hedonic functions commonly violate standard assumptions of uncorrelated error terms.

@@

One challenge and opportunity for researchers is to properly define the spatial extent at which to conduct their analysis.

Spatial heterogeneity

Spatial autocorrelation

 data collection and spatial modeling easier than in the past. However, the larger sample sizes and spatial extents create issues of heterogeneity as the relationship between characteristics and housing prices may differ across housing submarkets.   Day (2003) investigated the impact of traffic noise in Glasgow, Scotland and found four separate submarkets with varying relationships between noise and prices. Interestingly, traffic noise coefficients were estimated to be negative in three out of four submarkets, while the fourth revealed a significantly positive relationship. The positive relationship is hypothesized to be related to increased transportation access and mobility in an outer suburb of the city. In a similar study, Day et. al (2007) found traffic noise estimates to be significantly negative in five of eight submarkets using appriximately 10 thousand residential sales in 1997 in Birmingham, England.

@@

Theebe (2004) also used market segmentation to estimate more than 160,000 property sale values in the western part of the Netherlands sold between 1997 and 1999.  First, Theebe split the full sample into five parts according to political jurisdictions. Through spatial autoregression techniques, Theebe’s results indicated that sales values are unaffected by noise levels below 65 dB. Above 65 dB, NDSI is relatively small, ranging from 0.3 to 0.5\% per dB. An analysis of submarket data yielded larger results although these findings are less precise. While Theebe covered a large study area, the noise levels are obtained for square areas of 100 x 100m using the cumulative 24 hours energy level index Leq. This study uses a Leq index at a square area of 10 x 10m, therefore each property has a unique and continuous noise decibel measurement. Consequentially, the analysis on smaller submarket samples will be relatively more precise in terms of traffic noise estimates than Theebe (2004).

@@

\section{Data}
The 2010 US Census lists the population of the Minneapolis and St. Paul Metropolitan area in Minnesota as almost 3 million residents spread over seven counties. This study examines single family residential home transactions in the Census-defined urban areas of the three of the seven counties – Dakota, Ramsey and Washington County (see Figure 1). We obtained data from approximately forty thousand transactions between 2005 and 2010 (n=42,095) from the 2010 MetroGIS Regional Parcel Dataset published by the Metropolitan Council. In addition to the geographic location and date of the house sale, we collected structural and locational variables commonly used in the hedonic literature. See Table 1 for a complete list of the explanatory variables and their descriptive statistics.

@@

\subsection{Other Locational Attributes}
A common real estate addage states that the three most important things about real estate are: location, location and location. Knowing where each house is located allows us to also construct a vector of other attributes associated with the sales transaction. For instance, using GIS software we are able to calculate the Euclidean distance to numerous points of interest within the dataset, such as central business districts, shopping centers, parks, and lakes.

@@

\subsection{Time}
Given that we have data from before and after the recession of 2008-09, we subset the data in Table 2 by year of sale to look for differences. In addition to the noticeable decline in nominal sales values, there is also a drop in the number of sales across years. For example, in 2010 there were only 4,206 qualified property sales within the dataset, a more than 50\% reduction from 2005. While the mean sale price declined from roughly \$280,000 pre-crash to \$230,000 post-crash, most of the structural and neighborhood variables are relatively consistent across years. However, traffic noise displays a noticeable difference in the pre- and post-crash market transactions (the mean drops from 58 dB to 52 dB).

\section{Basic Econometric Model}\label{basicModel}
Consistent with past research, this study implements a standard semi-logarithmic hedonic pricing model. Our aim is to estimate the marginal willingness to pay for different attributes, in particular changes in the traffic noise associated with the house. Equation~\eqref{eq:model} expresses the basic hedonic model.
\begin{equation}\label{eq:model}	
ln \textrm{ Sale Price}_i = \beta _0 + \beta _1 \textrm{Noise}_i+ \beta _2 S_i+ \beta _3 N_i + \beta _4 T_i + \textrm{error}_i
\end{equation}
Where Noise$_i$ is the noise level for house $i$, $S_i$ is a vector of the house's structural attributes, $N_i$ is a vector of the neighborhood attributes, and $T_i$ is a vector of year fixed effects. We can interpret the regression model coefficients as the price semi-elasticities of the underlying attributes. For instance, we can interpret the coefficient on noise as the percentage increase in price for a one decibal increase in the traffic noise associated with the transaction in our dataset.

>>>>

\title{A Hedonic Price Analysis of Traffic Noise in the Twin Cities Housing Market}
\date{}
\begin{document}
\maketitle

@@

\section{Past Research}\label{sec:lit}
The value of noise pollution is typically estimated through simulated or related markets because no explicit private market for noise abatement exists. \citet{Nelson1982} is perhaps the best known earliest review of 14 studies of the economic impact of traffic noise on property values in Canada and the United States. More recently, a number of reviews have been produced for European government agencies. \citet{Bateman2001} reviewed the literature for the Scottish Government while \citet{Navrud2002} provided a ``state of the art'' review for the European Commission Environment Directorate General.

The Hedonic Price Method is commonly used to value noise by examining real estate price fluctuations as a function of noise levels. Based on the canonical work by \citet{Rosen1974}, researchers commonly estimate a hedonic price function specifying the logged sale price of a property as a linear function of the traffic noise exhibited at the house. Such a specification leads to reporting the impact of traffic noise as a percentage decrease in house price per one decibal (dB) increase in traffic noise. According to \citet{Batemen2001}, the negative effect of traffic noise has ranged from 0.08\% to 2.22\% per dB with a mean around 0.55 percent per dB. \citet{Navrud2002} found an average decrease of 0.64\% per dB in a review of 65 noise evaluation studies.

\citet{Nelson2008} produced an updated review of the traffic noise literature for the edited volume \underline{Hedonic Methods in Housing Markets Economics}. Nelson describes several areas of continued interest in the empirical literature, most notably: spatial modeling of the hedonic price function, long-run vs. short-run housing dynamics, and the appropriate measure of traffic noise. We now explain each of these areas in turn and our work's contribution to the field.

\subsection{Spatial Analysis}
Advances in Geographic Information System (GIS) technology and digitized datasets have made research easier to conduct over larger geographies, with larger samples, and while explicitly considering the spatial nature of the housing market. Estimation of the hedonic function using Ordinarly Least Squares (OLS) is typically inappropriate because the housing data commonly exhibit spatial autocorrelation. In particular, spatially autocorrelated error terms will lead to inconsistent estimates of OLS regression standard errors and inappropriate measures of statistical inference.

One common reason for significant spatially autocorrelated error terms is hedonic function misspecification. First, a portion of the hedonic regression errors may be explained by omitted spatial variables. Thus, nearby houses will tend to have similar spatial variable values and then similar error terms. Second, model misspecification can occur due to spatial non-stationarity of the hedonic function. That is, housing submarkets exist and the impact of changing housing characteristics will vary across submarkets. Failing to properly account for such heterogeneity will lead to spatially autocorrelated error estimates.

How do people do submarkets? 
\begin{itemize}
  \item Geography 
    \begin{itemize}
      \item census tracts
    \end{itemize}
  \item Political Boundaries
    \begin{itemize}
      \item city boundaries
    \end{itemize}
  \item Other (Data Mining)
\end{itemize}

@@

Day (2003) investigated the impact of traffic noise in Glasgow, Scotland and found four separate submarkets with varying relationships between noise and prices. Interestingly, traffic noise coefficients were estimated to be negative in three out of four submarkets, while the fourth revealed a significantly positive relationship. The positive relationship is hypothesized to be related to increased transportation access and mobility in an outer suburb of the city. In a similar study, Day et. al (2007) found traffic noise estimates to be significantly negative in five of eight submarkets using appriximately 10 thousand residential sales in 1997 in Birmingham, England.

@@

Theebe (2004) also used market segmentation to estimate more than 160,000 property sale values in the western part of the Netherlands sold between 1997 and 1999.  First, the data were split into five parts according to political jurisdictions. Through spatial autoregression techniques, Theebe’s results indicated that sales values are unaffected by noise levels below 65 dB. Above 65 dB, NDSI is relatively small, ranging from 0.3 to 0.5\% per dB. An analysis of submarket data yielded larger results although these findings are less precise. While Theebe covered a large study area, the noise levels are obtained for square areas of 100 x 100m using the cumulative 24 hours energy level index Leq. This study uses a Leq index at a square area of 10 x 10m, therefore each property has a unique and continuous noise decibel measurement. Consequentially, the analysis on smaller submarket samples will be relatively more precise in terms of traffic noise estimates than Theebe (2004).

@@

\subsection{Temporal Variation}
A common strategy is to analyze cross-sectional data. However, housing transactions take place at a specific point in time and the vast majority of data sets will contain some temporal variation. Some researchers may simply assume the hedonic function is constant thoughout their timespan. Others allow the function to shift up or down through the use of temporally varying intercept terms.

Few researchers allow the hedonic function to change over time. \citet{Cho} is a recent example of work allowing the hedonic regression coefficients to change over time, thereby estimating the changes in the implicit prices of hedonic characteristics over time.

\citet{Wilhelmsson2000} split his 290 Swedish house sales into two separate time periods: 1986-89 and 1990-95. A Chow test revealed that the impact of traffic noise was different in the earlier vs. later time period, with noise being considered more of a nuisance in the 1990s in the suburb of Stockholm.

We estimate regression coefficients using only sales within the past 12 months over the years 2005 - 2010. Such an analysis allows us to see if and when the implicit price of traffic noises changes within our study area.

\subsection{Measure of Traffic Noise}

\citet{HughsSirmans} use traffic volume.

Spatial resolution

\section{Data}
The 2010 US Census lists the population of the Minneapolis and St. Paul Metropolitan area in Minnesota as almost 3 million residents spread over seven counties. This study examines single family residential home transactions in the Census-defined urban areas of the three of the seven counties – Dakota, Ramsey and Washington County (see Figure 1). We obtained data from approximately forty thousand sales transactions between 2005 and 2010 (n=42,095) from the 2010 MetroGIS Regional Parcel Dataset published by the Metropolitan Council. In addition to the geographic location and date of the house sale, we collected structural and locational variables commonly used in the hedonic literature. See Table 1 for a complete list of the explanatory variables and their descriptive statistics.

@@

\subsection{Other Locational Attributes}
A common real estate addage states that the three most important things about real estate are: location, location and location. Knowing where each house is located allows us to also construct a vector of other attributes associated with the sales transaction. For instance, using GIS software we are able to calculate the Euclidean distance to numerous points of interest within the dataset, such as the nearest central business district, shopping centers, parks, and lakes.

@@

\subsection{Time}
Given that we have data from before and after the recession of 2008-09, we subset the data in Table 2 by year of sale to look for differences. In addition to the noticeable decline in nominal sales values, there is also a drop in the number of sales across years. For example, in 2010 there were only 4,206 property sales within the dataset, a more than 50\% reduction from 2005. While the mean sale price declined from roughly \$280,000 pre-crash to \$230,000 post-crash, most of the structural and neighborhood variables are relatively consistent across years. However, traffic noise displays a noticeable difference in the pre- and post-crash market transactions (the mean drops from 58 dB to 52 dB).

\section{Basic Econometric Model}\label{basicModel}
Consistent with past research, this study implements a semi-logarithmic hedonic pricing model. Our aim is to estimate the marginal willingness to pay for different attributes, in particular changes in the traffic noise associated with the house. Equation~\eqref{eq:model} expresses the basic hedonic model.
\begin{equation}\label{eq:model}	
ln \textrm{ Sale Price}_i = \beta _0 + \beta _1 \textrm{Noise}_i+ \beta _2 S_i+ \beta _3 N_i + \beta _4 T_i + \textrm{error}_i
\end{equation}
Where Noise$_i$ is the noise level for house $i$, $S_i$ is a vector of the house's structural attributes, $N_i$ is a vector of the neighborhood attributes, and $T_i$ is a vector of year fixed effects. We can interpret the regression model coefficients as the price semi-elasticities of the underlying attributes. For instance, we can interpret the coefficient on noise as the percentage increase in price for a one decibal increase in the traffic noise associated with the transaction in our dataset.
